The value of being indispensable: a cooperative approach to bargaining with a continuum of players

### 2023-11-22 src/paper/theory.tex
Start working on example application

\section{Many-sided case}

@@

>>>>

\section{Many-sided case}
\label{sec:many_sided}

@@

\section{Example application}

In this section I will apply the results of the previous sections to a simple model of two-sided platforms.
I follow the modelling approach of the monopoly platform case in \textcite{armstrong2006competition}, with a few extensions.
There are two major departures from the original model.
First, and most importantly, instead of letting the platform choose entry fees, I assume that they are the result of a bargaining process, and the outcomes are described by the weighted value.
Second, instead of assuming that each player's utility is linear in the number of the other side's players, I allow for more general (power) functions, which lets me capture the substitutability of the two sides.

\subsection{Model}

Imagine a two-sided market with a continuum of players on both sides.
Furthermore, assume that there is a single platform that connects the two sides.
The utility that a player on side $i$ derives from participating in the market is an increasing function\footnote{
    \textcite{armstrong2006competition} -- along with much of the literature on two-sided markets \parencite[e.g.][]{rochet2003platform,hagiu2006pricing} -- assumes that the utility of a player is linear in the number of players on the other side.
    Using a more general function allows me to model different degrees of network effects.
} of the number of players on the other side ($n_j$), minus the entry fee charged by the platform ($p_i$):
\begin{align*}
    u_i = \alpha_i n_j ^ {\gamma_i} - p_i,
\end{align*}
where $\alpha \geq 0$ and $\gamma > 0$ determine the strength and shape of the network effects, respectively.

Following \textcite{armstrong2006competition}, let us also model player entry in a reduced form manner.
Assume that there is a continuum of potential entrants on both sides, and the number of actual entrants is an increasing function of the utility achieved after entering\footnote{
    One possible microfoundation of this is assuming some idiosyncratic entry cost for the players.
    Then, the number of entrants is the number of players for whom the utility of entering is higher than the entry cost.
}:
\begin{align*}
    n_i = \phi_i(u_i).
\end{align*}

Finally, let us assume that the platform can charge lump-sum entry fees to both sides.
These fees can also be negative, in which case they are interpreted as subsidies.
The platform's profit is the sum of entry fees, minus the cost of serving the entrants ($F_i$ for each entrant):
\begin{align*}
    \pi(n_1, n_2) = n_1 p_1 + n_2 p_2 - F_1 n_1 - F_2 n_2.
\end{align*}

I will consider two cases as to the timing and nature of the price-setting process, as well as a welfare-maximizing benchmark.
For the latter, I derive the entry fees that maximize social welfare, i.e. the sum of the utilities of the two sides plus the platform's profit.
After that, I will first consider a model analogous to that in \textcite{armstrong2006competition}, where the platform can unilaterally commit to any entry fees it wishes.
Finally, I consider a model in which entry fees are the result of a bargaining process in such a way that the resulting utilities and profits correspond to the various players' weighted values.

\subsection{Benchmark: welfare-maximizing entry fees}

Let us consider first what entry fees would maximize social welfare.
Instead of approaching this problem directly, let us rely on the following observation: if entry fees correspond to the externality that a player's entry imposes on the other side, then the resulting choices are welfare-maximizing.
In this specific case, there are two externalities to consider.
First, it costs the platform $F_i$ to serve an additional player on side $i$.
Second, the entry of a player on side $i$ increases the utility on side $j$ due to positive network effects.
Welfare is maximized when the price reflects the balance of these two  factors:
\begin{align*}
    p_i^* &= F_i - n_j \frac{\partial u_j}{\partial n_i} \\
          &= F_i - \alpha_j \gamma_j n_j n_i^{\gamma_j - 1}.
\end{align*}
As in \textcite{armstrong2006competition}, for any $\alpha_j > 0$ welfare-maximizing $F_i$ is below the platform's cost of serving player $i$.
Thus, the platform's profit would necessarily be negative.
It naturally cannot happen in the case when the platform sets the entry fees, or even the bargaining case.

As an immediate corollary, utilities in this case are given by
\begin{align*}
    u_i^* = \alpha_i n_j ^ {\gamma_i} + \alpha_j \gamma_j n_j n_i^{\gamma_j - 1} - F_i.
\end{align*}
Furthermore, the utility of type $i$ depends positively on both $\alpha_i$ (mechanically) and $\alpha_j$ (a stronger positive externality on the other players implies a lower entry fee and thus higher utility).
Finally, if $n_i \geq e^{-\frac{1}{\gamma_j}}$, then $u_i$ is also increasing in $\gamma_j$.
Both of the latter two effects are related to player $i$ being compensated for the positive externality it imposes on the other side.

\subsection{Platform sets prices unilaterally}

Now let us examine what happens when the platform sets the entry fees unilaterally.
First, rewrite the platform's profit as a function of utilities, rather than the number of entrants:
\begin{align*}
    \pi(u_1, u_2) = \phi_1(u_1) [\alpha_1 \phi_2(u_2) ^ {\gamma_1} - u_1 - F_1] + \phi_2(u_2) [\alpha_2 \phi_1(u_1) ^ {\gamma_2} - u_2 - F_2].
\end{align*}

Let us assume that the above expression is concave, and thus the first order conditions characterize its maximum.
Then, simply taking the partial derivatives with respect to $u_i$ yields the following condition for the optimal entry fees.
\begin{proposition}
    \label{prop:platform_unilateral}
    If the platform can set entry fees unilaterally, then optimal entry fees for side $i$ are characterized by
    \begin{align*}
        p_i^u &= F_i - \alpha_j \gamma_j n_j n_i^{\gamma_j - 1} + \frac{\phi_1(u_1)}{\phi'_1(u_1)} \\
              &= p_i^* + \frac{\phi_1(u_1)}{\phi'_1(u_1)}.
    \end{align*}
\end{proposition}

As one would expect, profit-maximizing entry fees are higher than welfare-maximizing ones.
In addition to the comparative statics of the welfare-maximizing case, the platform's entry fees also depend on the elasticity of entry.
When it is low, the platform will set a high entry fee, as the loss of entrants is small.

For ease of comparison with the bargaining case, let us also derive the utilities in this case:
\begin{align*}
    u_i^u = \alpha_i n_j ^ {\gamma_i} + \alpha_j \gamma_j n_j n_i^{\gamma_j - 1} - F_i - \frac{\phi_1(u_1)}{\phi'_1(u_1)}.
\end{align*}
Clearly, as equilibrium utilities must be smaller than in the welfare-maximizing case, due to $\phi_i$ being increasing, the number of entrants is also lower.

\subsection{Bargaining}

Finally, let us consider the case when the platform and the players bargain over the entry fees.
To do this, let us impose some structure on the timing of the game.
The idea is that even though the platform cannot commit to a specific entry fee structure anymore, it can commit to a bargaining process.
Specifically, it is able to commit to a bargaining weight for each side and itself: $\lambda_P, \lambda_1, \lambda_2$.

Then, the game proceeds as follows.
In the first period, the platform chooses $\lambda_P, \lambda_1$ and $\lambda_2$.
After that, potential antrants on both sides simultaneously decide whether to enter or not.
Finally, in the last period, the platform and the entrants bargain over the entry fees, and utilities are realized according to their weighted values.

Let us start by looking at the last period.
Assume that $n_1$ and $n_2$ players have decided to enter on sides $1$ and $2$, respectively.
Furthermore, assume that bargaining weights are $\lambda_1, \lambda_2$ and $\lambda_P$.

The characteristic function of the game takes a similar form to that in \cref{sec:many_sided}.
If the platform is not part of the coalition, then the value is zero.
Otherwise, it is simply the sum of the utilities of the two sides minus the cost of serving them.\footnote{
    The entry fees do not matter, as they are simply payments between coalition members, and thus do not affect the total value.
}
This value, as a function of the number of entrants, is
\begin{align*}
    f(n_1, n_2) = n_1 \alpha_1 n_2^{\gamma_1} + n_2 \alpha_2 n_1^{\gamma_2} - F_1 n_1 - F_2 n_2.
\end{align*}

The share received by each type of player is then easily established using \cref{prop:many_sided_weighted}.
\begin{proposition}
    \label{prop:platform_bargaining_last_period}
    \begin{align*}
        \pi_P(n_1, n_2) &= \int_0^1 \lambda_P t^{\lambda_P - 1} f(t_1^{\lambda_1} n_1, t_1^{\lambda_2} n_2) \dt \\
        u_i(n_1, n_2) &= \int_0^1 t^{\lambda_P} \partial_i f(t_1^{\lambda_1} n_1, t_1^{\lambda_2} n_2) \dt
    \end{align*}
\end{proposition}
